 

### 2021-05-24 vertical-thermal-review.Rmd
bit of rearranging

@@

**Biophysical drivers reviewed above influence leaf temperature $T_{leaf}$ distribution along the vertical gradient.**
Along the biophysical gradients, leaves adjust to their environment by maintaining, when possible, a $T_{leaf}$ that is optimal for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can be cooler than air at high temperature and warmer than air at low temperature (limited homeothermy) or can be drastically warmer than $T_{air}$ (megathermy). In some cases, $T_{leaf}$ can be equalled to $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles  [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. `r fig_leaf_T`) where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig. 2; @michaletzPlantThermoregulationEnergetics2015].

@@

This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. During $gs$, evapotranspiration facilitates heat dissipation. But when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].

Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].

@@

Leaf interiors are $RH$ saturated [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$. Whereas when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissapation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].

>>>>

**The biophysical drivers reviewed above influence leaf temperature $T_{leaf}$ distribution along the vertical gradient.**
Along the biophysical gradients, leaves adjust to their environment by maintaining, when possible, a $T_{leaf}$ that is optimal for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can be cooler than air at high $T_{air}$ and warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy). 
Often, however, $T_{leaf}$ remains similar to $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016 @michaletzPlantThermoregulationEnergetics2015].
This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. During $gs$, evapotranspiration facilitates heat dissipation. But when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].

@@

@@

Leaf interiors are $RH$ saturated [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$. Whereas when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissipation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
